Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-19 paper.tex
Updates from Yan

\begin{abstract}
Recombination is an important and ongoing feature of SARS-CoV-2 evolution,
and presents major challenges to [things important to public health].
Detection of new recombinant strains is challenging, with methods
[being a bit rubbish].
In addition, many basic approaches fundamental to the study of viral
evolution assume that recombination is negligible, in that a single
phylogenetic tree can represent the genetic ancestry of the
circulating strains. Here we present an initial version of
\texttt{sc2ts}, a method to automatically detect recombinants
in real-time and to fully integrate them into a cohesive joint
model of mutation, recombination and genetic inheritance known
as an Ancestral Recombination Graph (ARG). We infer two ARGs under
different sampling strategies, and study their properties.
One contains 1.27 million sequences
sampled up to June 30, 2021, and the second is more sparsely sampled,
consisting of 784K sequences sampled up to June 30, 2022.
We find that both ARGs are
% better words?
highly consistent with
known features of SARS-CoV-2 evolution, recovering the basic
backbone phylogeny, mutational spectra, and recapitulating
details on the majority of known recombinant lineages.
Using the well-established and feature-rich \texttt{tskit} library,
the ARGs can also be stored concisely and processed efficiently
using standard Python tools. For example, the ARG for 1.27 million
sequences---encoding the inferred reticulate ancestry,
genetic variation, and extensive metadata---requires
58MB of storage,
and loads in less than a second.
%JK Not sure if we'd actually say it like this, but this is the
% basic gist I'd like to get accross[
The ability to fully integrate the effects of recombination into
downstream analyses, to quickly and automatically detect new recombinants,
and to provide an efficient and convenient platform for computation
makes \texttt{sc2ts} a promising approach, worthy of
further investment and development.
\end{abstract}

\section{Introduction}
% Recombination is an important force in SARS-CoV-2, recombinants
% have arisen and they have spread to high frequencies
Recombination via template switching is a common feature
in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of more
immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
Early in the SARS-CoV-2 pandemic, the lack of genetic variation
meant that recombination had little effect.
In late 2020, the appearance and
spread of variants of concern (VoC), designated into classes such as Alpha and
Delta which harboured multiple characteristic mutations,
created the conditions required for recombination to occur
and recombinant strains with onward transmission were detected at
low frequency~\citep{Jackson2021-ik}.
More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for
recombination~\citep{Bal2022-hq}.
In early 2023, multiple recombinant lineages have successfully
established and spread to high frequency, and accounting for recombinant
ancestry is now essential in understanding the evolution of the SARS-CoV-2
virus.

% Recombination is hard to detect
Detecting recombination in SARS-CoV-2 is difficult, and identifying new
recombinant strains is a time-consuming, often manual process [Citation?].
Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of mutations from different lineages
\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz},
which can only identify inter-lineage recombination events specific to those motifs.
% % JK: I've left this out because I felt it broke the narrative up a bit, but
% % happy to put back in. (We're still citing the paper in a few places).
% Others have inferred recombinant
% genealogies to disentangle recombination from recurrent mutation, without
% having to pre-specify characteristic mutations for each lineage
% \citep{Ignatieva2021-rg}.
More recently, \cite{Turakhia2022-it} presented a
phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find
candidate recombinants, RIPPLES first builds a phylogeny without assuming
recombination and then scans for long branches (i.e., carrying many mutations)
and then determines if these candidates would be better explained by
recombination. Specifically, RIPPLES exhaustively breaks each candidate
sequence into segments and attempts to find more parsimonious placements for
each segment on the phylogeny. If such placements are found, the sequence is
identified as a putative recombinant. Although it enables rapid searches for
genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
accounts for only mutations, treats recombinants \textit{post hoc}, and is an
incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
As noted by the authors, a \textit{post hoc} treatment of recombination is
possible when recombinant lineages are rare and leave few descendants. However,
the proliferation of recombinant lineages is making this increasingly
untenable; for example, more than half of the sequences sampled in February 2023
are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
future evolution of SARS-CoV-2 is likely to involve multiple sequential
recombination events on top of existing recombinant lineages, creating a highly
reticulated ancestry.

% PARA 3: but you can't use a phylogeny when there's recomb. Need a joint
% model of mutation and recomb. ARGs, tskit, forward refs to ARG section for
% details.

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analysis, e.g., inference of
selection~\citep{Anisimova2003-vr}. Standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}, and there is
no standard method for incorporating the effects of recombination into
phylogenetic analysis.
Ancestral Recombination Graphs (ARGs) are a means of describing such
network-like ancestry~\citep{Griffiths1981-lw,Gusfield2014-qw}, but
until recently lacked software support and sufficiently scalable
inference methods to be of practical use.
However, approaches to infer ARGs now exist that can scale to tens of
thousands of human genomes and beyond
\citep{Speidel2019-yh,Kelleher2019-ba,Schaefer2021-yg,Zhang2023-lf}, and which
deal with levels of repeated recombination far in excess of those seen in viral
phylogenies. The ``succinct tree sequence'' is an ARG data structure
which has led to significant computational advances across a range
of
applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,Ralph2020-efficiently,
Wohns2022-th}, and the supporting \texttt{tskit} software library is now widely
is now widely used in population genetic applications.
The methods in \texttt{tskit} have been developed to support millions of
whole human genomes~\citep{Kelleher2019-ba}, and so it is particularly well suited
to representing large SARS-CoV-2 genealogies,
which currently encompasses over 15 million sequences in the GISAID database
\citep{Shu2017-hp}.
See Section~\ref{sec:args} for more details on ARGs and the succinct
tree sequence data structure.

@@

@@

@@

Table~\ref{tab:args} also summarises some of the computational
properties of the inferred ARGs.
The ARGs are encoded as a ``succinct tree sequence'' using
the \texttt{tskit} library, which provides an extensive
suite of operations for constructing and analysing ARGs
(Section~\ref{sec:args}). For example, the Wide ARG
which contains complete genomes (with imputed missing data)
for around 1.2 million samples, along with extensive sample
and debugging metadata, requires only 58MB of space (compressed
using the \texttt{tszip} utility). The majority of this
space is used by metadata, which when discarded results in
an encoding that requires an average of only 8.29 bytes per
SARS-CoV-2 genome stored.
Loading these ARGs takes less than a second, and they can be interactively
analysed using Jupyter notebooks~\citep{Kluyver2016-jupyter}
on a standard laptop. The majority
of the analyses in this preprint can be carried out in seconds
with the \texttt{tskit} Python API, using a few gigabytes of RAM.

@@

\subsection{Ancestral Recombination Graphs}
\label{sec:args}
The term ``Ancestral Recombination Graph'' was introduced by
Griffiths and colleagues~\citep{Griffiths1991-two,Griffiths1998-ancestral}
and originally defined as an alternative formulation of the coalescent
with recombination stochastic process~\citep{Hudson1983-properties}.
Subsequently, the term ARG came to be used in a more general way to
describe not just realisations of this model, but to any
recombinant genetic ancestry~\citep{Minichiello2006-mapping,Zhang2023-lf}.
While there is some subtlety in the details~\citep{Wong2023-efficient},
we can think of an ARG as being any graph that encodes the
reticulate genetic ancestry of a sample of sequences under
the influence of recombination.
This definition encompasses various types of graphs traditionally
described using the broader term of phylogenetic networks.

@@

The \texttt{tskit} library~\citep{Tskit2023-tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with ARGs. Based on core functionality written
in C, it provides interfaces in C, Python and Rust. The Python interface
is based on NumPy~\citep{Harris2020-array}, and provides a convenient
platform for interactive analysis of large-scale data using, for
example, Jupyter notebooks~\citep{Kluyver2016-jupyter} and taking
advantage of the analysis tools in the burgeoning PyData ecosystem.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.][]{Haller2019-slim,Speidel2019-yh,
Terasaki2021-geonomics,
Fan2022-genealogical,Korfmann2022-weak,
Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}.
This ecosystem for storing and manipulating ARGs can
generally be used to efficiently record and analyse SARS-CoV-2 genealogies reconstructed
using other methods, not only the new approach that we describe here.

>>>>

\begin{abstract}
Recombination is an important and ongoing feature of SARS-CoV-2 evolution,
and presents major challenges to [things important to public health].
Detection of new recombinant strains is challenging, with methods
[being a bit rubbish].
In addition, many basic approaches fundamental to the study of viral
evolution assume that recombination is negligible, in that a single
phylogenetic tree can represent the genetic ancestry of the
circulating strains. Here we present an initial version of
\texttt{sc2ts}, a method to automatically detect recombinants
in real-time and to fully integrate them into a cohesive joint
model of mutation, recombination and genetic inheritance known
as an Ancestral Recombination Graph (ARG). We infer two ARGs under
different sampling strategies, and study their properties.
One is densely sampled, and includes 1.27 million sequences
sampled up to June 30, 2021.
The second is more sparsely sampled,
consisting of 784K sequences sampled up to June 30, 2022.
We find that both ARGs are
% better words?
highly consistent with
known features of SARS-CoV-2 evolution, recovering the basic
backbone phylogeny, mutational spectra, and recapitulating
details on the majority of known recombinant lineages.
Using the well-established and featured-rich \texttt{tskit} library,
the ARGs can also be stored concisely and processed efficiently
using standard Python tools. For example, the ARG for 1.27 million
sequences---encoding the inferred reticulate ancestry,
full alignments, and extensive metadata---requires
58MB of storage,
and loads in less than a second.
%JK Not sure if we'd actually say it like this, but this is the
% basic gist I'd like to get accross[
The ability to fully integrate the effects of recombination into
downstream analyses, to quickly and automatically detect new recombinants,
and to provide an efficient and convenient platform for computation
makes \texttt{sc2ts} a promising approach, worthy of
further investment and development.
\end{abstract}

\section{Introduction}
% Recombination is an important force in SARS-CoV-2, recombinants
% have arisen and they have spread to high frequencies
Recombination via template switching is a common feature
in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of more
immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
The lack of genetic variation made recombination challenging to detect early in the SARS-CoV-2 pandemic. In late 2020, the appearance and
spread of variants of concern (VoC), designated into classes such as Alpha and
Delta which harboured multiple characteristic mutations, increased the
detectability of recombination.
More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for
recombination~\citep{Bal2022-hq}.
[FIXME say this better and add citation.]
Multiple recombinant lineages have successfully
established and spread to high frequency.

[PARA 2: recombination is hard to detect. Quick overview of state-of-the-art.
Finish with phylogeny-based method, RIPPLES.]

[PARA 3: but you can't use a phylogeny when there's recomb. Need a joint
model of mutation and recomb. ARGs, tskit, forward refs to ARG section for
details.]

@@

% By searching for genomes containing segments
% from different VoC designations, multiple recombinant lineages have been
% detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha
% and Epsilon \citep{Wertheim2022-hj}, and Alpha and
% Delta~\citep{Sekizuka2022-xz}.

% Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
% combine specific subsets of mutations from different lineages
% \citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
% This approach only identifies those inter-lineage recombination events
% resulting in the patterns investigated.
% Others have inferred recombinant
% genealogies to disentangle recombination from recurrent mutation, without
% having to pre-specify characteristic mutations for each lineage
% \citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it} presented a
% phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
% over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find
% candidate recombinants, RIPPLES first builds a phylogeny without assuming
% recombination and then scans for long branches (i.e., carrying many mutations)
% and then determines if these candidates would be better explained by
% recombination. Specifically, RIPPLES exhaustively breaks each candidate
% sequence into segments and attempts to find more parsimonious placements for
% each segment on the phylogeny. If such placements are found, the sequence is
% identified as a putative recombinant. Although it enables rapid searches for
% genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
% accounts for only mutations, treats recombinants \textit{post hoc}, and is an
% incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
% As noted by the authors, a \textit{post hoc} treatment of recombination is
% possible when recombinant lineages are rare and leave few descendants. However,
% this is becoming increasingly untenable with the proliferation of recombinant
% lineages; for example, more than half of the sequences sampled in February 2023
% are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
% future evolution of SARS-CoV-2 is likely to involve multiple sequential
% recombination events on top of existing recombinant lineages, creating a highly
% reticulated genealogy.

@@

% It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
% and affects the results of downstream analysis, e.g., inference of
% selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
% not account for recombination
% \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
% genetic ancestries of recombining lineages can be represented by ancestral
% recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
% capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
% development of the ``succinct tree sequence'' data
% structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
% encoding of large ARGs. This representation should be well-suited to storing
% large recombinant viral genealogies such as SARS-CoV-2,
% which currently encompasses over 15 million sequences in the GISAID database
% \citep{Shu2017-hp}). Moreover the \texttt{tskit} software library, which implements this
% data structure, enables efficient manipulation of the stored genealogies and is
% integrated with a suite of open-source software tools to simulate, analyse, and
% infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

@@

Table~\ref{tab:args} also summarises some of the computational
properties of the inferred ARGs.
The ARGs are encoded as a ``succinct tree sequence'' using
the \texttt{tskit} library, which provides an extensive
suite of operations for constructing and analysing ARGs
(Section~\ref{sec:args}). For example, the Wide ARG
which contains complete genomes (with imputed missing data)
for around 1.2 million samples, along with extensive sample
and debugging metadata, requires only 58MB of space (compressed
using the \texttt{tszip} utility). The majority of this
space is used by metadata, which when discarded results in
an encoding that requires an average of only 8.29 bytes per
SARS-CoV-2 genome stored.
Loading these ARGs takes less than a second, and they can be interactively
analysed using Jupyter notebooks on a standard laptop. The majority
of the analyses in this preprint can be carried out in seconds
with the \texttt{tskit} Python API, using a few gigabytes of RAM.

@@

\subsection{Ancestral Recombination Graphs}
\label{sec:args}
The term ``Ancestral Recombination Graph'' was introduced by
Griffiths and colleagues~\citep{Griffiths1991-two,Griffiths1998-ancestral}
and originally defined as an alternative formulation of the coalescent
with recombination stochastic process~\citep{Hudson1983-properties}.
Subsequently, the term ARG came to be used in a more general way to
describe not just realisations of this model, but to any
recombinant genetic ancestry~\citep{Minichiello2006-mapping}.
While there is some subtlety in the details~\citep{Wong2023-efficient},
we can think of an ARG as being any graph that encodes the
reticulate genetic ancestry of a sample of sequences under
the influence of recombination. This definition encompasses various types of graphs traditionally described using the broader term of phylogenetic networks.

@@

The \texttt{tskit} library~\citep{Tskit2023-tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with ARGs. Based on core functionality written
in C, it provides interfaces in C, Python and Rust. The Python interface
is based on NumPy~\citep{Harris2020-array}, and provides a convenient
platform for interactive analysis of large-scale data using, for
example, Jupyter notebooks~\citep{Kluyver2016-jupyter} and taking
advantage of the analysis tools in the burgeoning PyData ecosystem.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.][]{Haller2019-slim,Speidel2019-yh,
Terasaki2021-geonomics,
Fan2022-genealogical,Korfmann2022-weak,
Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}. We emphasise that this ecosystem for storing and manipulating ARGs can generally be used to efficiently record and analyse SARS-CoV-2 genealogies reconstructed using other methods, not only the new approach that we describe here.
